---
title: "How to use the impgrplasso package"
author: "Drew Day"
date: "12/03/2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use the impgrplasso package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(kableExtra)
```

# Introduction

Least Absolute Shrinkage and Selection Operator (LASSO) regression is a 
statistical technique to impose penalties on independent variables in a 
regression in order to produce a more sparse model to improve out-of-sample 
prediction as well as to enhance the interpretability of the model. The L1 penalty
of the LASSO model can shrink independent variables to zero, in effect removing 
them from regression models. There is no clear way of defining traditional 
frequentist error estimates (i.e., confidence intervals and p-values) for the 
coefficient estimates in a LASSO regression model since any null distribution
for a given independent variable would inevitably result in some models omitting
that variable and others not. For this same reason, traditional approaches for 
combining regression estimates from multiply imputed data do not apply to LASSO
regression.

Multiple imputation is a series of methods for imputing multiple potential values 
for a missing observation, often based on a probability distribution that is 
predicted by the other independent and dependent variables to be used in a 
regression incorporating that variable with missingness. Multiple imputation 
with chained equations (MICE) is a well-established multiple imputation method
that begins by imputing the mean for each missing value of each variable with 
missingness except for one, using the now "complete" data of other variables to
predict values using regression models for the missing values in that selected 
variable and then use those as imputed values in the next regression imputing 
values for the next variable, and so on in several iterations of chained 
regressions. For more explanation see [Azur et al. 2011](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3074241/).

The goal of this package is to adapt one of several suggested methods for 
combining multiple imputation and LASSO regression, namely the MI-LASSO method of
[Chen and Wang 2013](https://onlinelibrary.wiley.com/doi/10.1002/sim.5783), which stacks imputed data sets
and then uses a group LASSO penalty to jointly penalize each imputed version of
a given independent variable. The 'impgrplasso' package builds off of this method,
which provided the best combination of interpretability and low prediction error
in the comparative simulation study of [Gunn et al. 2023](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10117422/), by
incorporating the following additional features:

  - The use of a prediction loss measure for optimizing the lambda parameter as 
  suggested by Gunn et al. 2023 rather than using the Bayes Information 
  Criterion (BIC), adding a log-loss measure for LASSO logistic regression 
  prediction error to the mean squared prediction error (MSE) proposed by Gunn 
  et al. in the case of LASSO standard regression
  - The incorporation of k-fold cross validation to optimize lambda rather than 
  single training/test splits
  - The incorporation of the established [‘grplasso’](https://CRAN.R-project.org/package=grplasso) package ([Meier et al. 2008](https://rss.onlinelibrary.wiley.com/doi/10.1111/j.1467-9868.2007.00627.x)) 
  for computing group LASSO penalties
  - The option to not penalize a subset of independent variables in the LASSO regression
  - Automatic detection of dummy variables from the same categorical variable and
  the application of a group LASSO penalty to those variables

## About LASSO Regression

Standard LASSO regression minimizes the loss function below:

$$
\sum_{i=1}^{N}{(y_i-[\beta_0+\sum_{j=1}^{p}{\beta_j x_{ij}]})^2} + \lambda \sum_{j=1}^{p}{\left\lvert{\beta_j}\right\lvert}
$$

Notice how the formula to the left of the addition sign is just the function that 
an ordinary least squares (OLS) regression minimizes (i.e., the sum of the squared
residuals). The right of the addition sign is the penalization term, which is
the sum of the absolute values of all independent variable coefficients $\beta_j$
scaled by the parameter $\lambda$.

## About MI-LASSO

The Chen and Wang 2013 MI-LASSO approach to LASSO regression for multiply imputed
data minimizes the similar function below:

$$
\sum_{d=1}^{m}{\sum_{i=1}^{N}{(y_{di}-[\beta_{0d}+\sum_{j=1}^{p}{\beta_{dj} x_{dij}}])^2}} + \lambda \sum_{j=1}^{p}\sqrt{\sum_{d=1}^{m}{\beta^2_{dj}}}
$$

Here the loss function is being summed over $m$ imputed data sets. In practice, we
stack the imputed data sets into a model matrix in a diagonal fashion, such that 
each imputed version of a variable gets a unique column and there are columns of 
ones for each imputed data set-specific intercept. Then the rows are the 
observations repeated $m$ times with zero values for all but the relevant $d$th 
column for each variable. As an example, let's say I have 3 observations and 2 
variables ($X1$ and $X2$) with two imputed data sets. $X1$ is continuous, but 
$X2$ is categorical and has three levels, meaning that two dummy variables will 
be generated (the referent level is omitted). Using the vertical bar to denote
the imputed version of each variable and the underscore to denote levels of a
categorical variable, I will get the following model matrix:

```{r echo=F}
dat<-data.frame(ID=rep(paste0("ID",1:3),2),beta0.1=rep(c(1,0),each=3),
                X1.1=c(6.5,0.2,3.3,rep(0,3)),
                X2_Level1.1=c(1,0,1,rep(0,3)),X2_Level2.1=c(1,0,1,rep(0,3)),
                beta0.2=rep(c(0,1),each=3),X1.2=c(rep(0,3),5.9,0.2,3.1),
                X2_Level1.2=c(rep(0,3),1,1,0),X2_Level2.2=c(rep(0,3),1,0,1))
names(dat)<-gsub("\\.","|",names(dat))
kable(dat,"html")%>%
  kable_styling()
```

The parameter $\lambda$ has to be chosen a priori before each model run, and as
$\lambda$ increases, the degree of penalization increases. Using k-fold cross-
validation, a $\lambda$ value that minimizes out-of-sample prediction loss 
function can be determined. The range of possible $\lambda$ will be higher than 
for a typical LASSO regression because of the high number of imputed columns and 
the group LASSO penalty between them.

# How to use the `impgrplasso` package

To use `impgrplasso`, we must first generate a list of imputed data frames. This 
can be done for example using the `mice` function from the 'mice' package or 
using the function `mice_by_group` from this package. The list of data frames 
must have all categorical columns transformed into dummy variables. This can 
either be done beforehand or the user can set the argument 'dummify' to TRUE 
(the default), in which case the function will do that transformation. Note that 
for proper functioning, variable names should not contain underscores ("_") so 
that when the dummy variables are generated, an underscore can be added by 
`impgrplasso` to separate the variable name (before the added underscore) from 
its respective levels (after the added underscore).

In this example, we will use MICE and create a list of imputed data frames using
the `nhanes` example data frame included in the 'mice' package.

```{r}
suppressPackageStartupMessages(library(mice))
nh <- nhanes
nh$hyp <- as.factor(nh$hyp - 1)
# above brings "hyp" from a {1, 2} set of unique values to {0, 1}

imp <- mice(nh, printFlag = FALSE)
dlist <- lapply(1:imp$m, function(x) complete(imp, x))
```

## mice_by_group

Another option is to use the function `mice_by_group` to generate group-specific
imputed data, for example for imputing within study cohorts or study sites in a 
multi-cohort study.

### Arguments

* `Data`: A data frame containing all variables (columns) to be used in 
the MICE equations and a column with the grouping variable.
* `groupvar`: The column name in `Data` for the grouping variable. If
this column is not of the class 'factor', it will be converted to a factor
with the default order of factor levels.
* `miceArgs`: These are arguments to be passed on to the `mice` function from 
the 'mice' package. These include the number of imputed data frames to be 
generated `m` (default = 10), the maximum number of MICE iterations `maxit` 
(default = 10), a random seed to be set for the MICE for reproducibility `seed` 
(default = 123), and a Boolean option for verbose printing `printFlag` (default 
= FALSE). All other `mice` function arguments default to the same values as in 
the 'mice' package (see `?mice::mice` for details). Note that the default `m` and 
`maxit` for `mice` are both 5, though I set them to 10 for this function for 
more iterations and better convergence, respectively.  

### Outputs

The output of `mice_by_group` is a list of imputed data frames.

### Examples

```{r}
library(mice)
library(impgrplasso)

nh$Group <- as.factor(c(rep(LETTERS[1:2], times = c(6, 7)), 
                        rep(LETTERS[1:2], times = c(6, 6))))

dlist2 <-  mice_by_group(nh, "Group", list(m = 5, maxit = 5, seed = 2, 
                                           printFlag = FALSE))
```

## impgrplasso

Now that we have a list of imputed data frames, we can use impgrplasso.

### Arguments

 * `impdatlist`: A list of data frames generated by multiple imputation.
 * `lams`: Either a single lambda value or a vector of lambda values. If one
 value is provided, the function will perform the MI-LASSO on the full data
 using only that single lambda value. If a vector of values is provided, the
 function will perform cross-validation to obtain fold- and imputation-specific
 as well as averaged loss values and coefficients for each lambda value so 
 that an optimal lambda can be chosen.
 * `outname`: A column name for the dependent variable. This can be continuous
 or binary. In the case of the latter, a LASSO logistic regression will be 
 performed.
 * `prednames`: Column names for the independent variables. These should be
 free of underscores. If dummy variables have already been made for the list
 of data frames `impdatlist`, provide those names with underscores between
 the variable name and the level (e.g., "sex_male"). 
 * `forcedin`: This is an optional vector of column names that should be a 
 subset of `prednames` that identifies unpenalized variables to be
 "forced in" to the model due to them not experiencing any shrinkage.
 * `kfolds`: If a vector of lambda values is provided for the argument `lams`, 
 `kfolds` provides the number of cross-validation folds for
 the cross-validation process. This defaults to 10.
 * `kfoldseed`: The rows chosen for each k-fold cross-validation are randomly
 selected once at the beginning of the function and then this selection is used
 consistently throughout subsequent steps. This value sets a seed for that 
 random process. This defaults to 10.
 * `scalecenter`: If TRUE (default), the function will scale and center all
 variables prior to calculations by subtracting the means of each variable and 
 dividing by the standard deviations. This is recommended for LASSO regression,
 but one can also perform their own form of standardization in the list of data
 frames and set this to FALSE if they prefer another standardization approach.
 * `dummify`: If TRUE (default), the function will produce dummy variables
 for all variables in `prednames` of classes 'factor' or 'character'. This
 can be set to FALSE if one wants to create dummy variables ahead of time and 
 include them in the list of data frames `impdatlist`.

### Outputs

`impgrplasso` outputs a list of class 'impgrplasso' with the following items:

 * `Coef`: If `lams` is a vector, this is a list of fold-specific data frames 
 containing coefficients for each variable for specific imputed data frames and 
 lambda values. If `lams` is a single value, this is a single data frame with 
 variable- and imputation-specific coefficients.
 * `MeanCoef`: If `lams` is a vector of length > 1, this is a data frame of mean 
 coefficients across imputed data frames for each variable at each fold and 
 lambda value. If `lams` is a single value, this is a data frame with mean 
 coefficients across imputed data frames for each variable.
 * `Loss`: This appears only if `lams` is a vector of length > 1, and it is a 
 data frame showing fold- and imputation-specific loss values and means for each 
 value of lambda.
 * `MeanLoss`: This appears only if `lams` is a vector of length > 1, and it is 
 a data frame showing the mean prediction loss for each lambda value.
 * `Model`: This appears only if `lams` is a single value, and it is the grplasso 
 model object run on the full data.
 * `Index`: This is the index argument for the grplasso function generated by 
 impgrplasso.
 * `allX`: This appears only if `lams` is a single value, and it is the large 
 stacked matrix of independent variables across imputed data frames generated by 
 `impgrplasso`.

A summary method is available for this output, either providing the average 
prediction loss per lambda value if multiple lambda values were input or the 
LASSO regression coefficients averaged over imputed data sets if one lambda value
was input.

### Examples

```{r}

# Searching for the best lambda

multiple_milasso <- impgrplasso(impdatlist = dlist, lams = seq(1, 5, 2), 
outname = "hyp", prednames = c("age", "bmi", "chl"), forcedin = "age", 
kfolds = 4, kfoldseed = 12)

summary(multiple_milasso)
```
```{r}
# Single Run

singlerun_milasso <- impgrplasso(dlist, 3, "hyp", c("age", "bmi", "chl"), "age")
summary(singlerun_milasso)

```
